---
title: "T-Mobile Homeruns"
author: "Evan Howard"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(tidyverse)
library(baseballr)
library(dplyr)
library(ggpubr)
library(formattable)
library(reshape)

SEA <- read.csv("SEAballparkHR.csv")
```


```{r, echo = FALSE}
x <- c(-.8, 1.2, 1.2, -.8, -.8)
z <- c(1.5, 1.5, 3.5, 3.5, 1.5)
sz <- tibble(x, z)

#visualizing pitch spray chart of all HR @ SEA with pitch_name
SEAstrikezone <- ggplot()+
  geom_path(data = sz, aes(x = x, y = z))+
  coord_equal()+
  labs(title = "HR at T-Mobile by Pitch Type")+
  xlab(NULL)+
  ylab("Feet Above the Ground")+
  geom_point(data = SEA, aes(x = plate_x,y = plate_z, color=pitch_name),
             na.rm = TRUE)+
  theme_pubr(base_size = 12, base_family = 'serif', legend = "right")

SEAstrikezone
```

Above is the strikezone plot of all the homeruns from our data file by pitch type.


```{r}

#plotting heat map for HR in seattle
heatmapSEA <- ggplot(SEA, aes(plate_x, plate_z))+
  stat_density_2d_filled(geom = "density_2d_filled", contour = TRUE, 
                         contour_var = "density", 
                         na.rm = TRUE, show.legend = TRUE)+
  geom_path(data = sz, aes(x, z), color = 'white')+
  xlab(NULL)+
  ylab("Feet Above Ground")+
  labs(title = "Home Run Density in Strikezone at T-Mobile", 
       fill = "Density Value")+
  theme_pubr(base_size = 12, base_family = 'serif', legend = "right")

heatmapSEA
```

Above is the heatmap to see the density of HR in relation to where they're located
in the strikezone at T-Mobile.

```{r}
plot_data <- ggplot_build(heatmapSEA)$data[[1]]

#finding HR that have are in the density <= 0.1
filtered_events <- plot_data %>% filter(piece == 1) %>%
  mutate(
    x= round(x,2),
    y= round(y,2))

#finding the events whose density <= 0.1 based off pitch location in SEA table
table_data <- SEA %>%
  filter(plate_x %in% filtered_events$x & plate_z %in% filtered_events$y)

#viewing events whose density is <= .1
ggplot()+
  geom_path(data = sz, aes(x = x, y = z))+
  coord_equal()+
  labs(title = "HR at T-Mobile by Pitch Type")+
  xlab(NULL)+
  ylab("Feet Above the Ground")+
  geom_point(data = table_data, aes(x = plate_x,y = plate_z, color=pitch_name),
             na.rm = TRUE)+
  theme_pubr(base_size = 12, base_family = 'serif', legend = "right")
```
Above we are viewing the pitches that were hit for home runs at T-Mobile park
whose density value was less than or equal to 0.1.

```{r}
ggplot(table_data, aes(x = reorder(pitch_name, -table(pitch_name)[pitch_name]),
                       fill = pitch_name))+
  geom_bar()+
  xlab("Pitch Name")+
  ylab("Count") +
  ggtitle("Count of Pitches Hit For HR Whose Density <= 0.1")+
  theme(text = element_text(family = 'serif', size = 14),
        axis.text.x = element_text(angle = 45, hjust = 1), 
        panel.background = element_blank())+
  guides(fill = 'none')
```


Above are the count of pitch types that were hit for HR at T-Mobile park
whose density value was less than or equal to 0.1.

```{r}
#viewing barrel contact of HR whose density was <= 0.1
ggplot(table_data, aes(x = factor(launch_speed_angle)))+
  geom_bar(na.rm = TRUE)+
  scale_x_discrete(labels = c("Under", "Flare/Burner", "Solid Contact", "Barrel"),
                   na.translate = FALSE)+
  xlab("Categorization Barrel Contact")+
  ylab("Count") +
  ggtitle("Categorization of Barrel Contact by Count")+
  theme(text = element_text(family = 'serif', size = 14), 
        panel.background = element_blank())

```


Above we are viewing the barrel contact of HR at T-Mobile park whose density
value was less than or equal to 0.1.





